Hierarchical recurrent models reveal latent states of neural activity in \textit{C. elegans}

### 2017-12-20 drafts/elife/draft.tex
adding figures for elife version of paper
some text for figure 1

\documentclass{article}

@@

\subsection*{Hierarchical models pool information while allowing individual variability}
% Introduce the probabilistic model

\begin{figure}[t]
\centering%
\includegraphics[width=6in]{figures/figure1} 
\caption{A schematic of the hierarchical model for combining
  information across multiple worms in order to learn a shared
  dynamical system for \celegans.  \textbf{a.}~All worms share a
  canonical dynamics library. Here, we have three dynamics functions
  (blue, red, gold), each of which is a simple linear mapping from
  $x_t \in \reals^2$ to $x_{t+1} \in \reals^2$.  The recurrent model
  assigns probabilities to the different dynamics functions based on
  the current location in $x$-space.  \textbf{b.}~The worms also share
  an emission matrix, $C$, which can be viewed as an embedding of
  neurons in latent space.  \textbf{c.}~Each worm has its own dynamics
  library, which allows for slight deviations from the canonical
  dynamics.  \textbf{d.}~The model segments neural activity into a
  sequence of discrete states~$z_{t} \in \{1, 2, 3\}$, which index
  into the dynamics library.  \textbf{e.}~The discrete states specify
  the dynamics that are used to propagate the continuous latent
  states~$x_t$ forward in time. Here, the two dimensional latent
  states are colored according to the instantaneous discrete
  state. \textbf{f.}~The observed neural activity is a linear
  projection of the underlying continuous latent states.  On any given
  trial, we observe only a fraction of all the neurons.  The
  hierarchical model combines these partial recordings to infer the
  underlying states, the dynamics libraries, and the shared neural
  embedding. }
\label{fig:hslds}
\end{figure}

@@

\subsection*{AR-HMM's automatically parse data into discrete states with simple dynamics}

@@

@@

>>>>

\documentclass[11pt]{article}

@@

\subsection*{Hierarchical state space models combine information across
  trials to learn canonical dynamics}

% Goal: learn dynamical model of neural activity
% Challenge 1: how to capture nonlinear dynamics of neural activity with simple model? 
% Solution 1: recurrent state space models
% Challenge 2: how to combine information across trials?
% Solution 2: hierarchical model

We seek a parsimonious characterization of the dynamics of neural
activity in the head ganglia of \celegans. Recent advances in optical
imaging offer simultaneous recordings of most of the head ganglia
neurons, but we must surmount two main challenges in order to extract
a dynamics model from these data.  We need to design a model that is
flexible enough to capture nonlinear dynamics yet interpretable enough
to reveal simplifying structure when it exists. Moreover, our model
must accept partial recordings from separate organisms to learn shared
neural dynamics yet still allow for individual variability. We
construct a hierarchical state space model that satisfies both
criteria, decomposing complex dynamics into a collection of simpler
pieces that can be learned from noisy, partial recordings with
trial-to-trial variability.

We study the data presented in~\citet{kato2015global} and
\citet{nichols2017global}, which consist of a collection of optical
recordings of calcium fluorescence in immobilized \celegans. The
recordings are 18 minutes long with a frame rate of about 3Hz.  On
average, about 100 single units are extracted from each
recording. \celegans~is genetically stereotyped and each of its 302
neurons has a unique label (e.g. \textsf{AVAL}). On average, 30 single
units could be unambiguously labeled in each recording.

To model this data, we build on switching linear dynamical systems
(SLDS) \citep{chang1978state, ackerson1970state, hamilton1990analysis,
  ghahramani1996switching, murphy1998switching}, a type of state space
model that views data as a projection of a low-dimensional latent
state that switches between linear dynamical regimes.  Formally,
let~$y_t \in \reals^N$ denote the vector of neural activity observed
at time~$t$.  The expected activity is given by,~$\E[y_t] = g(x_t)$,
where~$x_t \in \reals^D$ is a continuous latent state
and~$g: \reals^D \to \reals^N$ is a linear map. We assume~$D \ll N$,
reflecting the low-dimensional nature of neural activity. SLDS
simplify the dynamics of the continuous latent states by introducing a
discrete latent state~$z_t \in \{1, \ldots, K\}$ that indicates the
current dynamical regime.  Each discrete regime~$k$ is associated with
a linear map~$f_k: \reals^D \to \reals^D$. The discrete state~$z_t$
specifies which map to use in order to propagate~$x_t$ to the next
time step.  Specifically,~$\E[x_{t+1}] = f_{z_t}(x_t)$. By switching
between different linear dynamical regimes, the SLDS captures highly
nonlinear dynamics.  We seek to infer the continuous and discrete
latent states and simultaneously learn the linear maps, the
distribution of the noise in the observations and the continuous
states, and the dynamics that govern how the discrete latent states
evolve.

% Recurrent model

% Robust model

% Hierarchical model

% Describe figure 1

\begin{figure}[t]
\centering%
\includegraphics[width=6in]{figures/figure1} 
\caption{A schematic of the hierarchical model for combining
  information across multiple worms in order to learn a shared
  dynamical system for \celegans.  \textbf{a.}~All worms share a
  canonical dynamics library. Here, we have three dynamics functions
  (blue, red, gold), each of which is a simple linear mapping from
  $x_t \in \reals^2$ to $x_{t+1} \in \reals^2$.  The recurrent model
  assigns probabilities to the different dynamics functions based on
  the current location in $x$-space.  \textbf{b.}~The worms also share
  an emission matrix, $C$, which can be viewed as an embedding of
  neurons in latent space.  \textbf{c.}~Each worm has its own dynamics
  library, which allows for slight deviations from the canonical
  dynamics.  \textbf{d.}~The model segments neural activity into a
  sequence of discrete states~$z_{t} \in \{1, 2, 3\}$, which index
  into the dynamics library.  \textbf{e.}~The discrete states specify
  the dynamics that are used to propagate the continuous latent
  states~$x_t$ forward in time. Here, the two dimensional latent
  states are colored according to the instantaneous discrete
  state. \textbf{f.}~The observed neural activity is a linear
  projection of the underlying continuous latent states.  On any given
  trial, we observe only a fraction of all the neurons.  The
  hierarchical model combines these partial recordings to infer the
  underlying states, the dynamics libraries, and the shared neural
  embedding. }
\label{fig:model}
\end{figure}

@@

\begin{figure}[t]
\centering%
\includegraphics[width=6in]{figures/figure2} 
\caption{}
\label{fig:emissions}
\end{figure}

\subsection*{SLDS automatically parse data into discrete states with simple dynamics}

\begin{figure}[t]
\centering%
\includegraphics[width=6in]{figures/figure3} 
\caption{}
\label{fig:syllables}
\end{figure}

@@

\begin{figure}[t]
\centering%
\includegraphics[width=6in]{figures/figure4} 
\caption{}
\label{fig:recurrent}
\end{figure}

@@

\begin{figure}[t]
\centering%
\includegraphics[width=6in]{figures/figure5} 
\caption{}
\label{fig:o2}
\end{figure}
